---
title: "Population Genetics CSP Report"
author: "DRAG1 Downstream Pipeline"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_name: "unknown_run"
  report_dir: ""
  data_dir: ""
  figures_dir: ""
  tables_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(ggplot2)

run_name <- params$run_name
data_dir <- params$data_dir
figures_dir <- params$figures_dir
tables_dir <- params$tables_dir
```

## Run Summary

```{r}
run_summary <- read_tsv(file.path(tables_dir, "csp_report_run_summary.tsv"), show_col_types = FALSE)
kable(run_summary)
```

## Haplotype Diversity and Frequencies

```{r}
metrics <- read_tsv(file.path(tables_dir, "csp_haplotype_diversity_metrics.tsv"), show_col_types = FALSE)
kable(metrics)
```

```{r}
freq_group <- read_tsv(file.path(tables_dir, "csp_combined_haplotype_frequency_by_group.tsv"), show_col_types = FALSE)
kable(head(freq_group, 30))
```

## TH2R/TH3R Haplotype Network

```{r}
net_file <- file.path(figures_dir, "csp_th2r_th3r_haplotype_network.png")
knitr::include_graphics(net_file)
```

## Variant Matrix and Site Catalog

```{r}
site_catalog <- read_tsv(file.path(data_dir, "csp_window_site_catalog.tsv"), show_col_types = FALSE)
kable(site_catalog)
```

```{r}
hap_profile <- read_tsv(file.path(data_dir, "csp_haplotype_profiles.tsv"), show_col_types = FALSE)
kable(head(hap_profile, 25))
```

## Population Structure and Diversity Plots

### MAF Distribution

```{r}
maf_fig <- file.path(figures_dir, "csp_maf_distribution_by_group.png")
if (file.exists(maf_fig)) knitr::include_graphics(maf_fig)
```

### PCA (PC2 vs PC1)

```{r}
pca21 <- file.path(figures_dir, "pca_pc2_vs_pc1.png")
if (file.exists(pca21)) knitr::include_graphics(pca21)
```

### PCA (PC4 vs PC1)

```{r}
pca41 <- file.path(figures_dir, "pca_pc4_vs_pc1.png")
if (file.exists(pca41)) knitr::include_graphics(pca41)
```

### Nucleotide Diversity (Pi)

```{r}
pi_fig <- file.path(figures_dir, "nucleotide_diversity_pi_by_group.png")
if (file.exists(pi_fig)) knitr::include_graphics(pi_fig)
```

### Tajima's D

```{r}
tj_fig <- file.path(figures_dir, "tajimas_d_by_group.png")
if (file.exists(tj_fig)) knitr::include_graphics(tj_fig)
```

### Pairwise Fst

```{r}
fst_fig <- file.path(figures_dir, "pairwise_fst_distribution.png")
if (file.exists(fst_fig)) knitr::include_graphics(fst_fig)
```

## Methods and Evidence

This report uses SNP haplotypes in CSP TH2R and TH3R windows directly from merged cohort VCF genotypes, with population grouping from metadata-driven region assignment.

- Haplotype construction: TH2R (nt 940-988) and TH3R (nt 1030-1099) SNP windows were extracted from the CSP contig in merged VCF; per-sample alleles were encoded from genotype calls (`0/0` = reference allele, any non-zero genotype = alternate allele, missing = `NA`).
- Haplotype diversity was computed from per-group haplotype frequencies using unbiased sample-size correction.
- Network inference uses `pegas::haplotype` and `pegas::haploNet`; if diversity is insufficient, a diagnostic figure is produced instead of an empty plot.
- Population metrics and structure follow standard malaria population-genetics practice: MAF distribution, PCA, nucleotide diversity (Pi), Tajima's D, and pairwise Fst.

### Recent Supporting Literature (<=10 years)

- Dwivedi et al., 2020. *Population genetic analysis of PfCSP in two ecological regions in Ghana*. Malaria Journal. <https://malariajournal.biomedcentral.com/articles/10.1186/s12936-020-03510-3>
- Lyimo et al., 2024. *Signatures of balancing selection in the C-terminal region of CSP in Tanzania*. Malaria Journal. <https://malariajournal.biomedcentral.com/counter/pdf/10.1186/s12936-024-04974-3.pdf>
- Khan et al., 2023. *Genetic variations of PfCSP and implications for malaria vaccine efficacy*. Infection, Genetics and Evolution. <https://www.sciencedirect.com/science/article/pii/S1567134823000163>
- Chaudhury et al., 2021. *Breadth of humoral responses to CSP C-terminus and RTS,S efficacy*. Vaccine. <https://www.sciencedirect.com/science/article/pii/S0264410X20316352>
- Casares et al., 2022. *PfCSP C-terminal epitope/structural variation context relevant to immunity*. PLOS Pathogens. <https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1010409>

## Reproducibility Notes

- Report data: `05_report/data/`
- Report figures: `05_report/figures/`
- Report tables: `05_report/tables/`

This report artifact is intended to be regenerated by pipeline scripts, not edited manually.
